A National Pipeline System as a Solution to Future Water Shortage in the United States

### 2013-02-04 paper.tex
hey

Due to the scale and complexity of the task of water management in
the United States, we make several simplifying assumptions in order
to obtain a workable model.
\begin{itemize}
\item Counties are thought of as discrete entities, occupying points in
space located at their centroid. This assumption enables us to think
of counties as vertices (as in a graph). This assumption can be made
without losing too much accuracy for several reasons: The United States
is split into more than 3,000 counties, with an average area of less
than 3,300 square kilometers, and average length across of less than
60 kilometers. The level of inaccuracy is sufficiently low that it
is workable (in contrast, states for instance are far too large to
be accurately modeled as points).
\item De-salinization plants, reservoirs, and purification plants are all facilities that are built at a single location, and have the intended effect of providing more water to that location. For the purposes of our modelling, we think of them as being essentially the same kind of thing.
\item When considering the costs of pipelines, only the distance of pipeline
is considered. The landscape of a region may in practice affect the
length of pipeline required, or may result in additional cost per
length of pipeline, but we shall not model this, as we could not find
data on the difficulty of pipeline construction across different types
of terrain.
\item Pipelines are thought of as bi-directional. This assumption essentially
reflects the reality of pipelines. There are a few exceptional cases
where it is significantly more difficult to send water in one direction,
but these are rare enough that they can be disregarded.
\item Pipelines have variable rates of water transmission, depending on
factors such as size. In theory it would be possible to save money on pipelines by constructing smaller pipelines in areas where water rarely needs to be transported at high rates. 
However, the amount of water that needs to be transported through any given pipe is highly variable (ie. after a sudden intense rainfall). Thus, pipelines for long-distance water transmissions are relatively stable in cost. Our estimate below for pipeline cost is a typical figure.
\item Some pipelines exist already, but we do not consider them in our model. This is partially because they are likely to have minimal impact because there are not many of them, and partially because already-existing pipelines are poorly documented, so there is no way to efficiently collect and parse such data
\end{itemize}

@@

The cost of a de-salinization facility in 2013 is approximately \$14
per gallon per day\cite{key-12}.

@@

Edges in the Minimal Spanning Tree in Figure \ref{fig:graph} are weighted inversely proportional to cost and directly proportional to the populations at each of the endpoints, and sorted. The resulting sorted table gives a full list of all $3108$ pipelines sorted by this utility metric. The final (highest priority) rows are shown in Table \ref{table:priority}.

\begin{table}[h]\label{table:priority}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
FIPS 1 & FIPS 2 & Priority \tabularnewline
\hline
48201 & 48339 & 10346.3 \tabularnewline
\hline
24005 & 24510 & 10675.4 \tabularnewline
\hline
6001  &6013 & 11019.8 \tabularnewline
\hline
34003 & 34017 & 11190.7 \tabularnewline
\hline
51153 & 51685 & 12104.5 \tabularnewline
\hline
11001 & 51013 & 12113.6 \tabularnewline
\hline
48029 & 48091 & 13787.1 \tabularnewline
\hline
34017 & 36061 & 16584.3 \tabularnewline
\hline
51059 & 51600 & 18528.8 \tabularnewline
\hline
51153 & 51683 & 22134.2 \tabularnewline
\hline
6037 & 6059  &23742.4 \tabularnewline
\hline
36047 & 36061  &27446.1 \tabularnewline
\hline
36059 & 36081  &29256.5 \tabularnewline
\hline
36005 & 36061 & 32595 \tabularnewline
\hline
36047 & 36081 & 33517.6 \tabularnewline
\hline
17031 & 17043 & 41637.4 \tabularnewline
\hline

@@

\subsection*{Computational Software}
All data analysis and visualization was performed using the technical computing language Julia, making use of the packages "DataFrames", "Gadfly", "Distributions", "Compose", "Graphs" all of which are available in the Julia package manager and were licensed under the MIT license.

@@

\subsubsection{Data Processing}
The dataset we obtained was in the proprietary, binary "Shapefile" format, that is used by many professional GIS systems. Luckily, there exists an open documentation of the format \cite{key-21}, which allowed us to implement our own parser to extract the locations, and shapes of the various counties.

@@

As outlined in Section 3, we made use various data points collected by the USGS (United States Geological Survey) for every single county in the United  States in the years 1985, 1990, 1995, 2000 and 2005. The data is split into 104 categories ranging from standard data points such as the total amount of water consumed in a specific specific county to more unexpected ones such as water irrigation due to golf courses.

@@

@@

        \begin{subfigure}[b]{0.4\textwidth}
                \centering
                \includegraphics[width=\textwidth]{95.png}
                \caption{1995}
                \label{fig:1995}
        \end{subfigure}
       \begin{subfigure}[b]{0.4\textwidth}
                \centering
                \includegraphics[width=\textwidth]{00.png}
                \caption{2000}
                \label{fig:2000}
        \end{subfigure}
        \caption{USGS Data on total water consumption in the indicated year. The more red, the higher the water consumption of the county.}\label{fig:water}
\end{figure}

For this data, we wrote a visualization software that allows arbitrary sets of county-level data to be displayed on maps of the United States. The visualization of total water usage for every county in the years covered by the USGS data is shown in figure \ref{fig:water}

@@

>>>>

Due to the scale and complexity of the task of water management in
the United States, we make several simplifying assumptions in order
to obtain a workable model.
\begin{itemize}
\item Counties are thought of as discrete entities, occupying points in
space located at their centroid. This assumption enables us to think
of counties as vertices (as in a graph). This assumption can be made
without losing too much accuracy for several reasons: The United States
is split into more than 3,000 counties, with an average area of less
than 3,300 square kilometers, and average length across of less than
60 kilometers. The level of inaccuracy is sufficiently low that it
is workable (in contrast, states for instance are far too large to
be accurately modeled as points).
\item De-salinization plants, reservoirs, and purification plants are all facilities that are built at a single location, and have the intended effect of providing more water to that location. For the purposes of our modelling, we can think of them as being essentially the same structure, with the proviso that they each can only be constructed in particular parts of the country. 
\item When considering the costs of pipelines, only the distance of pipeline
is considered. The landscape of a region may in practice affect the
length of pipeline required, or may result in additional cost per
length of pipeline, but we shall not model this, as we could not find
data on the difficulty of pipeline construction across different types
of terrain.
\item Pipelines are thought of as bi-directional. This assumption essentially
reflects the reality of pipelines. There are a few exceptional cases
where it is significantly more difficult to send water in one direction,
but these are rare enough that they can be disregarded.
\item Pipelines have variable rates of water transmission, depending on
factors such as size. In theory it would be possible to save money on pipelines by constructing smaller pipelines in areas where water rarely needs to be transported at high rates. 
However, the amount of water that needs to be transported through any given pipe is highly variable (ie. after a sudden intense rainfall). Thus, pipelines for long-distance water transmissions are relatively stable in cost. Our estimate below for pipeline cost is a typical figure.
\item Some pipelines exist already, but we do not consider them in our model. This is partially because they are likely to have minimal impact because there are not many of them, and partially because already-existing pipelines are poorly documented, so there is no way to efficiently collect and parse such data
\end{itemize}

@@

The cost of a de-salinization facility in 2013 is approximately \$14
per gallon per day\cite{key-12}. This figure is approximately the same as for similar kinds of facilities (reservoirs, dams).

@@

Edges in the Minimal Spanning Tree in Figure \ref{fig:graph} are weighted inversely proportional to cost and directly proportional to the populations at each of the endpoints, and sorted. The resulting sorted table gives a full list of all $3108$ pipelines sorted by this utility metric. The final (highest priority) rows are shown in Table \ref{table:priority}, with FIPS1 and FIPS2 representing the FIPS codes of the ends of each pipe.

\begin{table}[h]\label{table:priority}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
FIPS1 & FIPS 2 & Priority \tabularnewline
\hline
48201 & 48339 & 10346.3 \tabularnewline
\hline
24005 & 24510 & 10675.4 \tabularnewline
\hline
6001  &6013 & 11019.8 \tabularnewline
\hline
34003 & 34017 & 11190.7 \tabularnewline
\hline
51153 & 51685 & 12104.5 \tabularnewline
\hline
11001 & 51013 & 12113.6 \tabularnewline
\hline
48029 & 48091 & 13787.1 \tabularnewline
\hline
34017 & 36061 & 16584.3 \tabularnewline
\hline
51059 & 51600 & 18528.8 \tabularnewline
\hline
51153 & 51683 & 22134.2 \tabularnewline
\hline
6037 & 6059  &23742.4 \tabularnewline
\hline
36047 & 36061  &27446.1 \tabularnewline
\hline
36059 & 36081  &29256.5 \tabularnewline
\hline
36005 & 36061 & 32595 \tabularnewline
\hline
36047 & 36081 & 33517.6 \tabularnewline
\hline
17031 & 17043 & 41637.4 \tabularnewline
\hline

@@

\subsection*{Computational Software}
All data analysis and visualization was performed using the technical computing language Julia \cite{julia}, making use of the packages "DataFrames", "Gadfly", "Distributions", "Compose", "Graphs" all of which are available in the Julia package manager and were licensed under the MIT license.

@@

\subsubsection{Data Processing}
The dataset we obtained was in the proprietary, binary "Shapefile" format, that is used by many professional GIS systems. Luckily, there exists an open documentation of the format \cite{shapefile-gis}, which allowed us to implement our own parser to extract the locations, and shapes of the various counties.

@@

As outlined in Section 3, we made use various data points collected by the USGS (United States Geological Survey) for every single county in the United  States in the years 1985, 1990, 1995, 2000 and 2005. The data is split into 104 categories ranging from standard data points such as the total amount of water consumed in a specific specific county to more unexpected ones such as water irrigation due to golf courses. In addition to the actual data, every row also includes meta information such as the State in which the county is located and the FIPS (Federal Information Processing Standard) code of the county for which the data was collected.

@@

\subsubsection{Data Processing}
The standard format of the data allowed for an easy import into our computational software using the Julia "DataFrames" package. This allowed us to treat the data essentially as an im-memory database which we could query not only for the downloaded data, but which also allowed for more complex models to be executed on the data with relative ease. Treating the data as a database also tremendously helped with combining all the data from the many different sources. Not all of the data we obtained contained the same information as to their location, but after normalizing all the different kinds of information available to a FIPS code based primary key, standard database operations such as joining two datasets by primary key. This operation was incredibly helpful during visualization as we could just join the DataFrame containing the geographical information with that containing e.g. the water usage information and have the results be immediately

@@

        \begin{subfigure}[b]{0.4\textwidth}
                \centering
                \includegraphics[width=\textwidth]{95.png}
                \caption{1995}
                \label{fig:1995}
        \end{subfigure}
       \begin{subfigure}[b]{0.4\textwidth}
                \centering
                \includegraphics[width=\textwidth]{00.png}
                \caption{2000}
                \label{fig:2000}
        \end{subfigure}
        \caption{USGS Data on total water consumption in the indicated year. Values are represent on a logarithmic scale with green indicating values towards the bottom of the scale and red indicating values towards the top of the scale.}\label{fig:water}
\end{figure}

In order to be able to visualize the input data and any simulation results obtained from the models used in this paper, we created a custom visualization software based on the Julia "Compose" package, a modern implementation of Wilkinson-Wickham's Grammar of Graphics.

The visualization software can visualize arbitrary county level data sets as long as there is some way to identify  which county the data belongs to. This is easiest is the data in need of visualization is already keyed on the FIPS code of the county, but it is also possible to plot data that only contains longitudinal and latitudinal information of any point in the county, a feature which is very useful to visualize data coming from individual measurement stations within the county rather than values that are determined for the county as a whole.

We extended our software to be able to graph arbitrary directed and undirected Graphs (represented by the Julia "Graphs") between counties our points within these counties. This capability was used to generate the graph of the proposed pipeline network.

Our software also makes use of the vector graphics capabilities of the Compose framework to create almost infinitely scalable graphics that allowed us to visualize smaller areas very effectively. The figures contained within this document are rasterizations of these vector images  to ensure a reasonable file size.

@@

\section{Potential Weaknesses}

Our proposed pipeline network can play a huge role in managing the distribution of freshwater over the United States. However, it is not a perfect solution to water management problems, and has some potential flaws:
\begin{itemize}
\item Before completion, there are holes in the network: Before every pipeline has been constructed, 
there will be some pairs of counties that are not connected by pipelines. In this case, the network is still 
useful, but is not quite as powerful as a complete network would be.
\item Large scale droughts can be an issue: if the entirety of the one region (say, the west coast) is in 
drought, while (say) the east coast has a water surplus, there may be bottlenecks in the middle of the country 
as large amounts of water are transferred. This could potentially be an issue, but in reality, this kind of situation 
is extremely rare, partially because the coasts of the United States do not tend to be where droughts occur.
\item Some nearby counties are only connected by long and convoluted paths: This is not as much of an issue as 
one might imagine, because generally droughts happen slowly, and hence it is acceptable for water to take a long time to 
reach its destination. However, if this is seen as a major issue, a small number of additional pipes can be added on a case-by-case basis.
\end{itemize}

However, none of these is a fatal flaw in the idea of constructing a pipeline network.
